A Bayesian Framework for Learning Regular Expressions

### 2012-12-08 paper/paper.tex
added paper.tex
Format for regexes uses verbatim. Format for strings uses single quotes.

Below, we define the prior probability such that our model biases smaller hypotheses, and the likelihood of generating a given hypothesis is inversely proportional to its cost. Our model aims to balance generality with specificity. For example, given \verb!abc, abd!, we wish to favor the regex \verb!ab(c|d)! over \verb!A*! where \verb!A! represents any letter,\footnote{See the Appendix for a brief summary of our regex notation.} which is less likely given the size principle. At the same time, given \verb!aba!, \verb!abc!, \verb!abd!, \verb!abe!, \verb!abf!, we may wish to favor \verb!abA! over \verb!ab(a|c|d|e|f)!.

@@

Further, we define the probability with which a DFA would stop generating a string at a start state . For instance, given the regular expression \verb!a*!, we would need to define the probability of generating strings \verb!a!, \verb!aa!, \verb!aaa!, etc.

@@

% TODO: explain character scope!

\subsection{Generating Hypotheses}
The hypothesis space is initiated with the base hypothesis, the term applied to a DFA that accepts strictly the input strings. We begin generating a series of consecutive states, where the transition between the first state and the second state is governed by the first letter of the first string, the transition between the second and the third is governend by the the second letter, and so on. For each additional string, if the first i characters overlap with those of any of the first i characters of one previous string but differ in the $i+1$th, we add no new states to the DFA until the transition between the i-1th state and the ith state. We then generate new states for the remaining characters of the string. See Figure~\ref{base_hypothesis} for an example of the first hypothesis generated from a set $S = $\{\verb!ababb!, \verb!abbab!\} .
  \begin{figure}[ht]
\begin{center}
\begin{align*}
\includegraphics[scale=0.4]{readin.png}
\end{align*}
\end{center}
\caption{An example base hypothesis DFA generated from the strings `ababb' and `abbab'. Note that the two strings share the first two states, but that the fifth and tenth are distinct.} 
\label{base_hypothesis}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\text{(a)}\includegraphics[scale=0.4]{merge-0.png}
&\text{(b)}\includegraphics[scale=0.4]{merge-1.png}
&\text{(c)}\includegraphics[scale=0.4]{merge-2.png}
\end{align*}
\end{center}
\caption{An example of state merging. The DFA starts at state 0 and accepts only at the shaded states. The base hypothesis accepts ``b'' and ``ab'' (a). After merging states 0 and 2, the next DFA accepts an infinite set of strings (b). Finally, the merge of states 0 and 1 forces the merge of state 0 and 3 to maintain one unique outbound transition for any state and character (c).} 
\label{merge-figure}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\text{(a)}\includegraphics[scale=0.4]{wildcard-0.png}
&\text{(b)}\includegraphics[scale=0.4]{wildcard-1.png}
\end{align*}
\end{center}
\caption{An example of inserting a wildcard, ``A'', which represents any character. Replacing a wildcard for one of the outbound transitions of state 0 has forced a merge of states 1 and 3.} 
\label{wildcard-figure}
\end{figure}

@@

An upper bound for the complexity of model merging is much more extreme. Performing subset construction for an NFA with $n$ states can produce, in the worst case, a DFA with $2^n$ states. Assuming that any subset of these states also produces a valid DFA, we can have at most $2^{2^n-1}$ DFA.
% \footnote{See the Appendix for detailed sizing of the hypothesis space.} Actually, this seems inaccurate.

We resolve this issue by performing beam search. In beam first, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability. (For a more detailed discussion of Beam Search, see \cite{HwangStuhlmullerGoodman2011}

@@

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta(r_i, a_{i+1})) &= \frac{1}{\sum_{a\in A} 1\{\delta(r_i, a)\in Q\}}
\end{align*}
where $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta(r_i, a)\in Q$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard \verb!A!, the transition is defined for all 26 letters, so the denominator is 26.

For the cases in which a string may alternatively be accepted or continue, one must assign a probability for accepting or continuing the string. For instance, consider the DFA in part (c) of Figure~\ref{merge-figure}. The DFA may generate the empty string, \verb!E!, or it may generate any arbitrary string with \verb!a! and \verb!b!, like \verb!ababbba!. We define the probability of stopping at a given string to be a fixed value $\tau$. This makes shorter strings more likely to be generated.

@@

\section{Experiment Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string `fit' with a list of strings. A sample question was as follows: \\
\begin{center}Given the list:

@@

\subsection{Parameter Fitting}
Because the role of $\beta$ had the effect of biasing models with fewer transitions, it also had a tendency to bias regular expressions with fewer states. For instance, the $\beta$ term would prefer the regular expression \verb!a*! over \verb!aaaa!. Thus, this term was fitted before fitting the $\alpha$ term.

@@

% TODO: move this?
\begin{table}[!ht]
\begin{center} 
\caption{Parameter fitting experiments} 
\label{parameter-table} 
\vskip 0.12in
\begin{tabular}{llll} 
\hline
Input  & Test & Subject Accept & Model Accept
% \\ \hline
% bhi, bhw, bhi, & bho & 30\% & $\beta < 14$
% \\ bhw, bhw
\\ \hline
efg, hfg, ifg & ffg & 45\% & $\beta > 3$
\\ \hline
glts, elts, glts, & ilts & 70\% & $\beta > 5$
\\ flts
\\ \hline
ab, abb & abbb & 90\% & $\alpha = 0, \beta = 4$
\\ \hline
ao, aoo, aooo & aoooo & 80\% & $\alpha = 0, \beta=4$
\\ \hline
\end{tabular} 
\end{center} 
\end{table}

@@

We note that in cases 1 (S = \{\verb!yyyt!, \verb!yt!, \verb!yyyt!, \verb!yyyt!, \verb!yt!\}, s= \verb!yyyyyt!), 4 (S = \{\verb!jojojojo!, \verb!jojojojo!, \verb!jojojojo!\}, s = \verb!jojo!), and 6 (S = \{\verb!twtwtw!, \verb!twtwtw!, \verb!twtwtw!\}, s =  \verb!tw!), it is clear that the model weighs a smaller number of transitions greater than human subjects do. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases. This suggests that our model's parameter $\beta$ may in fact be better represented by more than one variable in human decision making.

@@

% \begin{table}[!ht]
% \begin{center} 
% \caption{Sample table title.} 
% \label{sample-table} 
% \vskip 0.12in
% \begin{tabular}{ll} 
% \hline
% Error type    &  Example \\
% \hline
% Take smaller        &   63 - 44 = 21 \\
% Always borrow~~~~   &   96 - 42 = 34 \\
% 0 - N = N           &   70 - 47 = 37 \\
% 0 - N = 0           &   70 - 47 = 30 \\
% \hline
% \end{tabular} 
% \end{center} 
% \end{table}

@@

\bibliographystyle{apacite}
\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
 \bibliography{regexbib}

\section{Appendix: Technical Notes}
\subsection{Regex and DFA Notation}
We used a limited alphabet wherein normal characters are all lower case. We also used the following rules to define regular expressions.
\begin{enumerate}
	\item Acceptance of specific strings is represented as concatenated characters. E.g. \verb!abc! accepts only \verb!abc!.
	\item Acceptance of disjunctive strings is represnted by \verb![a|b]!. E.g. \verb![abc|c]!accepts only \verb!abc! and \verb!c!.
	\item Acceptance of any character is represented by the wildcard \verb!A!. E.g. \verb!aA! accepts \verb!aa!, \verb!ab!, \verb!ac!, etc.
	\item Acceptance of the empty string is represented by \verb!E!. \verb!Ea! is the same as \verb!a!.
	\item Acceptance of any sequence repeated zero or more times is represented by the Kleene star, \verb!a*!. E.g. \verb!a*! accepts \verb!E!, \verb!a!, \verb!aa!, etc, 
\end{enumerate}

>>>>

Below, we define the prior probability such that our model biases smaller hypotheses, and the likelihood of generating a given hypothesis is inversely proportional to its cost. Our model aims to balance generality with specificity. For example, given `abc, abd', we wish to favor the regex \verb!ab[c|d]! over \verb!A*! where \verb!A! represents any letter,\footnote{See the Appendix for a brief summary of our regex notation.} which is less likely given the size principle. At the same time, given `aba', `abc', `abd', `abe', `abf', we may wish to favor \verb!abA! over \verb!ab[a|c|d|e|f]!.

@@

Further, we define the probability with which a DFA would stop generating a string at a start state . For instance, given the regular expression \verb!a*!, we would need to define the probability of generating strings `', `a', `aa', `aa', etc.

@@

% TODO: explain character scope'

\subsection{Generating Hypotheses}
The hypothesis space is initiated with the base hypothesis, the term applied to a DFA that accepts strictly the input strings. We begin generating a series of consecutive states, where the transition between the first state and the second state is governed by the first letter of the first string, the transition between the second and the third is governend by the the second letter, and so on. For each additional string, if the first i characters overlap with those of any of the first i characters of one previous string but differ in the $i+1$th, we add no new states to the DFA until the transition between the i-1th state and the ith state. We then generate new states for the remaining characters of the string. See Figure~\ref{base_hypothesis} for an example of the first hypothesis generated from a set $S = $\{`ababb', `abbab'\} .
  \begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.4]{base_hypothesis.png}
\end{center}
\caption{An example base hypothesis DFA generated from the strings `ababb' and `abbab'. Note that the two strings share the first two states, but that the fifth and tenth are distinct.} 
\label{base_hypothesis}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\text{(a)}\includegraphics[scale=0.4]{merge-0.png}
&\text{(b)}\includegraphics[scale=0.4]{merge-1.png}
&\text{(c)}\includegraphics[scale=0.4]{merge-2.png}
\end{align*}
\end{center}
\caption{An example of state merging. The DFA starts at state 0 and accepts only at the shaded states. The base hypothesis accepts `b' and `ab' (a). After merging states 0 and 2, the next DFA accepts an infinite set of strings (b). Finally, the merge of states 0 and 1 forces the merge of state 0 and 3 to maintain one unique outbound transition for any state and character (c).} 
\label{merge-figure}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
\begin{align*}
\text{(a)}\includegraphics[scale=0.4]{wildcard-0.png}
&\text{(b)}\includegraphics[scale=0.4]{wildcard-1.png}
\end{align*}
\end{center}
\caption{An example of inserting a wildcard, `A', which represents any character. Replacing a wildcard for one of the outbound transitions of state 0 has forced a merge of states 1 and 3.} 
\label{wildcard-figure}
\end{figure}

@@

An upper bound for the complexity of model merging is much more extreme. Performing subset construction for an NFA with $n$ states can produce, in the worst case, a DFA with $2^n$ states. Assuming that any subset of these states also produces a valid DFA, we can have at most $2^{2^n-1}$ DFA. \footnote{See the Appendix for detailed sizing of the hypothesis space.}

We resolve this issue by performing beam search. In beam search, we perform breadth first search, but for any given depth of the search tree, we only consider children of the $m$ most promising hypotheses. In our case, the most promising hypotheses were defined as those with the highest posterior probability. (For a more detailed discussion of Beam Search, see \cite{HwangStuhlmullerGoodman2011}.

@@

To simplify inference, we assume uniform probability over all possible outbound transitions for a given state. Formally, the probability of making a state transition is given by:
\begin{align*}
	p(\delta(r_i, a_{i+1})) &= \frac{1}{\sum_{a\in A} 1\{\delta(r_i, a)\in Q\}}
\end{align*}
where $1\{\cdot\}$ is the indicator function, which evaluates to $1$ if the enclosed proposition is true and $0$ otherwise. $\delta(r_i, a)\in Q$ indicates that the state transition leads to another state of the DFA, so the denominator counts the number of characters for which a transition is defined. In the case of the wildcard `A', the transition is defined for all 26 letters, so the denominator is 26.

For the cases in which a string may alternatively be accepted or continue, one must assign a probability for accepting or continuing the string. For instance, consider the DFA in part (c) of Figure~\ref{merge-figure}. The DFA may generate the empty string, `E', or it may generate any arbitrary string with `a' and `b', like `ababbba'. We define the probability of stopping at a given string to be a fixed value $\tau$. This makes shorter strings more likely to be generated.

@@

\section{Experiment Design and Model Fitting}
Having generated the above Bayesian model for selection of the regular expression with the highest posterior probability, we then conducted a behavioral experiment.  41 subjects (Median age 34, std-dev 13.3) were surveyed on Amazon Mechanical Turk. Two groups (one of 20, one of 21) were presented with 10 yes or no questions in which they were asked to determine whether a given string `fit' with a list of strings. A sample question was as follows:
\begin{center}Given the list:

@@

\subsection{Parameter Fitting}
Because the role of $\beta$ had the effect of biasing models with fewer transitions, it also had a tendency to bias regular expressions with fewer states. For instance, the $\beta$ term would prefer the regular expression `a*' over `aaaa'. Thus, this term was fitted before fitting the $\alpha$ term.

@@

% TODO: move this?
\begin{table}['ht]
\begin{center} 
\caption{Parameter fitting experiments} 
\label{parameter-table} 
\vskip 0.12in
\begin{tabular}{llll} 
\hline
Input  & Test & Subject Accept & Model Accept
% \\ \hline
% bhi, bhw, bhi, & bho & 30\% & $\beta < 14$
% \\ bhw, bhw
\\ \hline
efg, hfg, ifg & ffg & 45\% & $\beta > 3$
\\ \hline
glts, elts, glts, & ilts & 70\% & $\beta > 5$
\\ flts
\\ \hline
ab, abb & abbb & 90\% & $\alpha = 0, \beta = 4$
\\ \hline
ao, aoo, aooo & aoooo & 80\% & $\alpha = 0, \beta=4$
\\ \hline
\end{tabular} 
\end{center} 
\end{table}

@@

We note that in cases 1 (S = \{`yyyt', `yt', `yyyt', `yyyt', `yt'\}, s= `yyyyyt'), 4 (S = \{`jojojojo', `jojojojo', `jojojojo'\}, s = `jojo'), and 6 (S = \{`twtwtw', `twtwtw', `twtwtw'\}, s =  `tw'), it is clear that the model weighs a smaller number of transitions greater than human subjects do. However, because this same parameter in part controls for wildcards, lowering it reduces performance on many other cases. This suggests that our model's parameter $\beta$ may in fact be better represented by more than one variable in human decision making.

@@

% \begin{table}['ht]
% \begin{center} 
% \caption{Sample table title.} 
% \label{sample-table} 
% \vskip 0.12in
% \begin{tabular}{ll} 
% \hline
% Error type    &  Example \\
% \hline
% Take smaller        &   63 - 44 = 21 \\
% Always borrow~~~~   &   96 - 42 = 34 \\
% 0 - N = N           &   70 - 47 = 37 \\
% 0 - N = 0           &   70 - 47 = 30 \\
% \hline
% \end{tabular} 
% \end{center} 
% \end{table}

@@

\bibliographystyle{apacite}
\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
 \bibliography{regbib}

\section{Appendix: Technical Notes}
\subsection{Regex and DFA Notation}
We used a limited alphabet wherein normal characters are all lower case. We also used the following rules to define regular expressions.
\begin{enumerate}
	\item Acceptance of specific strings is represented as concatenated characters. E.g. \verb!abc! accepts only `abc'.
	\item Acceptance of disjunctive strings is represnted by \verb![a|b]!. E.g. \verb![abc|c]! accepts only `abc' and `c'.
	\item Acceptance of any character is represented by the wildcard \verb!A!. E.g. \verb!aA! accepts `aa', `ab', `ac', etc.
	\item Acceptance of any sequence repeated zero or more times is represented by the Kleene star, \verb!*!. E.g. \verb!a*! accepts `' (the empty string), `a', `aa', etc. 
\end{enumerate}
